Ensemble Averaged Patterns and Long-term Trends of Acceleration of Tropical Cyclones -- A Proxy For Extratropical Interactions

### 2020-11-05 paper/manuscript.tex
update

\title{Along-Track Acceleration of Tropical Cyclones: A Proxy For Extratropical Interactions}

@@

\begin{abstract}
  It is well known that rapid changes in tropical cyclone (TC) motion occur during interaction with extratropical wavetrains. While TC speed of motion, in general, has received much attention in the past, acceleration has not. Here, using a large data sample, we formally examine the synoptic-scale patterns associated with  tangential and normal components of TC acceleration. During rapid acceleration, a developing wavepacket in the extratropics is present and the TC ?????? with the TC straddled by an upstream trough and downstream ridge. This pattern is consistent with the process of extratropical transition (ET) of TCs. In contrast, during near-zero or rapid deceleration, the extratropical wavepacket remains poleward, with a ridge  located directly north of the storm center.  On average, the tangential acceleration peaks about 18 hours prior to the ET designation. The curvature acceleration increases before ET and stabilizes thereafter.
%
 We also find a robust decrease in the annually averaged values of rapid acceleration over the past 38 years within the latitude band most commonly associated with ET over the Atlantic. Consequently, we hypothesize that trends in rapid acceleration can be used as an independent metric to assess changes in the extratropical stormtrack activity that mediate ET.
\end{abstract}

@@

\introduction
The motion of a tropical cyclone (TC) is known to be  governed by the large-scale environment surrounding it \citep[e.g.][]{Hodanish1993}. As noted by \cite{E2018}, it was recognized early on that the forward speed of a TC can be closely approximated by the surrounding wind field.  It is also understood that a TC is not an isolated vortex that is passively carried by a current. The background environment, typically, also has attendant gradients of potential vorticity, moisture, and wind shear. As also summarized by \cite{E2018}, a TC actively responds to these external stimuli. As a consequence, not only its structure and intensity are impacted, the large-scale flow may also change significantly. This, in turn, may also affect its motion. An example of the response to background gradient of potential vorticity is the generation of the so called $\beta-gyres$, which contribute to TC motion. Similarly, vertical shear can modulate the intensity of the TC via ventilation of low entropy air into the storm core. This in turn can also impact TC motion. On the other hand, a TC can influence the large-scale flow by exciting baroclinic waves in the extratropical stormtrack, leading to rapid downstream development and rearrangement of the flow. Thus, it can be contended that a TC is always interacting with its environment and the interaction is manifested in its motion.

%

The aspects of TC motion that are particularly important for operational forecasts are track and forward speed. The track garners much attention for obvious reasons -- it  informs  potential locations that will be affected by a storm. TC forward speed is an important parameter that impacts storm surge, intensity changes and precipitation. There is a large body of literature dealing with various research problems related to TC tracks and speed. They are extensively documented in several recent reviews \citep[e.g.,][]{E2018,other}. In contrast, while forecasters have long known that TCs speed up rapidly while approaching an extratropical wave-train, relatively less attention has been devoted in the research literature to details of TC acceleration.

As noted by \cite{E2018}, steady progress has been made in our understanding of TC motion and this is reflected in our ability to forecast their tracks.

@@

\section{Data}
We used the IBTrACS v4 \citep{KnappIBTRaCS} data for track information. We did not apply a threshold for maximum sustained winds, but discarded all storms that lasted less than two days.  We focused on the years 1966--2019, spanning the \emph{satellite era} for the North Atlantic and Pacific basins (e.g.. Beven et al. 2003; Landea 2007; Vecchi and Knutson), for which the TC data is deemed relatively more reliable as compared to previous years. Selected atmospheric fields were taken from the the European Center for Medium Range Weather Forecasting's (ECMWF) ERA-Interim (ERAi) reanalysis \citep{DUS11} for the period 1981--2016.

 As our intent here was to examine the motion of tropical storms, in subsequent sections unless explicitly stated,  we focused on the track data prior to the completion of ET. To ascertain whether a tropical storm underwent ET, we used the \emph{nature} designation in the IBTraCS database that relies on the judgement of the forecasters from one or more agencies responsible for the ocean basin.  Admittedly, there will be some subjectivity in this designation. An alternative would be to employ a metric such as the cyclone phase space \citep{Hart2003}, usually calculated using meteorological fields from global numerical weather prediction models (e.g., reanalysis data). As we wish to be independent of modeled products to characterize the storm, we choose to use the forecaster designated storm nature. Furthermore, \cite{BCSEH2019a} found that the phase space calculation was sensitive to the choice of the reanalysis model. Occasionally, the nature of the storm is not recorded (NR) or, if there is an inconsistency among the agencies, it is designated as mixed (MX). In the north Atlantic, only a small fraction of track data ($\approx 0.5\%$) in the IBTrACs is designated as NR or MX. However, the fraction is higher in the North Pacific ($\approx 14\%$), and we return to this point in a subsequent section where it could introduce some uncertainty in our conclusion for this basin.

\section{Tangential and curvature acceleration}
The acceleration of a hypothetical fluid element moving with the center of a TC can be written as:
%
\begin{equation}
    \mathbf{a} = \frac{dV}{dt} \mathbf{\hat{s}} + \frac{V^2}{R} \mathbf{\hat{n}} 
    \label{eq:acc}
\end{equation}
%
where $V$ is the forward speed and $R$ is the radius of curvature of the track at a given location. Here, $\mathbf{\hat{s}}$ and $\mathbf{\hat{n}}$ are orthogonal unit vectors in the so-called \emph{natural coordinate} system. The former is directed along the TC motion. The latter is directed along the radius of curvature of the track.  The first term in equation \ref{eq:acc} is the tangential acceleration and the second is the curvature or normal acceleration. The speed $V_j$ at any track location (given by index j)  is calculated as:
%
\begin{equation}
V_j = \frac{(D_{j,j-1} +  D_{j,j+1})}{\delta t}
\end{equation}
%
where $D$ refers to the distance between the two consecutive points, indexed as shown above, along the track. Since we used 3-hourly reports from the IBtracs, $\delta t = 6$ h. The tangential acceleration was calculated from the speed using centered differences. To calculate the curvature acceleration, it is necessary to first calculate the radius of curvature, R. For this, we  fitted a circle at each track point using the method described in Appendix A. Figure \ref{fig:track} illustrates R at five selected locations along the the track of hurricane Katrina (2005).

%approximated as follows. The latitude-longitude positions of the TC track are projected to a 2-dimensional surface using the Mercator transformation. Then, a circle is fitted to each set of three consecutive track points denoted by $\mathbf{X}_{j-1}, \mathbf{X}_j, \mathbf{X}_{j+1}$. The resulting circle is used to assign the radius of curvature of the track point  $\mathbf{X}_j$.

@@

During 1966--2019, a total of 688 storms in the North Atlantic and 1676 storms in the western North Pacific passed the lifetime threshold of 2 days.

Figures \ref{fig:distNA} and \ref{fig:distWP} shows some basic statistics for TC speed and accelerations as a function of latitude. TC locations were sorted in $10^\circ$-wide latitude bins for this. Tables \ref{tab:AC1}-\ref{tab:AT2} provide the same data. The average speed of all North Atlantic storms (including over-land) is about 21 km/hr. The same for the western Pacific storms is about 19 km/hr. As expected, TC speed clearly varies with latitude. It is lower in the subtropics as compared to other latitudes. It increases sharply in the vicinity of $40^{o}$ N.

The tangential acceleration, as defined in eq. \ref{eq:acc}, can be positive or negative. The mean and median tangential acceleration in both basins remain near-zero equatorward of $30^\circ$N. This suggests that TCs in this region tend to translate steadily, or are equally likely to accelerate and decelerate. Poleward of $30^\circ$N, they are substantially positive, indicating that TCs in these latitudes are subject to rapid acceleration. For example, the mean tangential acceleration in the $30^\circ-40^\circ$N latitude band is about 5.4  km/hr day$^{-1}$ in the Atlantic and nearly 8  km/hr day$^{-1}$ in the Western Pacific.  The curvature acceleration, by our definition, takes only positive values and, and on average, steadily increases with latitude. The distribution of TC speed and tangential acceleration is relatively symmetric about the median value as compared to the curvature acceleration.

@@

We now examine the flow pattern associated with extremes in tangential and curvature acceleration of tropical storms. For this, storm-relative ensemble average fields are constructed as follows:

\begin{itemize}
    \item All TC track locations were binned into 10$^\circ$ wide latitude strips (e.g.,  10-20$^\circ$ N, 20-30$^\circ$ N, 30-40$^\circ$ N). For brevity, only results for the 30-40$^\circ$ N bin are discussed here.

    \item Taking all TC positions in each latitude bin, the mean ($\bar{A}$) and standard deviation ($\sigma$) of  acceleration (tangential and curvature separately) were calculated.

    \item The data in each bin were separated into three categories: (a) Rapid acceleration ($A > \bar A + s \sigma_A$); (b) Near Zero acceleration ($ |A - \bar{A}| < .025\sigma_A$); and (c) Rapid deceleration ($A  < \bar A - s\sigma_A$). We show results for s=1 here and note that they do not change qualitatively when we tried other values of s (=1.25, 1.5). TC positions that did not fall in these categories are discarded. Note: A TC could appear more than once within this latitude range at different times depending on its motion.

    \item Using the dates corresponding to each track position, ensemble average of large scale fields (e.g. geopotential) were calculated after shifting the grids such that the center of all storms are coincident. The centroid position of storms in that latitude range was used for the reference center and the reference time is denoted as Day-0.  This yields a storm-relative ensemble average view of the large scale atmospheric fields.

    \item  For curvature acceleration, which as defined here takes positive values, only the first two categories (rapid and near-zero) were considered.

@@

%------------------------------------------------------------------------------------------
\begin{figure*}[t]
  \centering
    \includegraphics[width=.95\textwidth, trim={0cm 8cm 0cm 4cm},clip]{composite_tang_40_30_NA_zero.pdf}
  \caption{Storm-relative composite average geopotential heights (thick solid lines) and anomalies (color shaded) for all TCs located in the latitude bin 30-40$^o$N over the Atlantic. The composite fields are shown at three levels -- 300 hPa, 500 hPa and 850 hPa. In each panel, the anomalous 1000 hPa geopotential is shown using thinner purple contours. Left, center and right columns show the fields for rapid acceleration, near-zero and rapid deceleration respectively.}
  \label{fig:lag0CompZ}
\end{figure*}

Figure \ref{fig:lag0CompZ} shows the composite geopotential heights (thick solid lines) and their anomalies (color shaded) calculated using all Atlantic TC locations within 30-40$^\circ$ N for the three categories of tangential acceleration. The anomalies were defined relative to a long term (1980--2016) climatology. The composite fields are shown at three levels -- 300 hPa, 500 hPa and 850 hPa. In each panel, the anomalous 1000 hPa geopotential is shown using thinner purple contours. It highlights the composite TC as well as the surface development within the stormtrack.

The ensemble average at reference time (Day=0) for rapid tangential acceleration cases (left column of Fig. \ref{fig:lag0CompZ}) shows the composite TC interacting with within a well defined extratropical wavepacket. The TC is straddled by an upstream trough and a downstream ridge. At 500 hPa, the geopotential anomalies of the TC and the upstream trough are clearly connected. The 1000 hPa geopotential anomaly field shows the near-surface cyclones and anticyclones of the extratropical stormtrack but is dominated by the cyclonic anomaly associated with the TC. The entire wavepacket shows upshear tilt of geopotential anomalies with height, indicating baroclinic growth. This arrangement of the TC and the extratropical wavepacket is consistent with the synoptic-scale flow that is typically associated with ET \citep[e.g.,][]{KHE2000,MGY2003,RJD2008,RJ2014,K2019}. We note that at this point, all storms in the ensemble average are still classified as \emph{tropical}. Thus, we interpret this composite as pre-ET completion state.

The ensemble average for near zero tangential acceleration cases (middle column of Fig. \ref{fig:lag0CompZ}) shows a very different synoptic scale pattern. The extratropical wavepacket is substantially poleward, with a ridge immediately north of the composite TC. The geopotential anomalies of the extratropical wavepacket and the composite TC appear to be distinct with no evidence of merger. Interestingly, the ensemble average for rapid tangential deceleration cases (right column of Fig. \ref{fig:lag0CompZ}) appears to be similar to the near-zero acceleration cases, albeit the downstream extratropical ridge is a little stronger in the latter.

%------------------------------------------------------------------------------------------
% 
To get a sense of the temporal evolution of the TC-extratropical wavepacket system, we now show the composites from Day-1 to Day+1 in Fig. \ref{fig:lagCompZ}. As in the previous figure, the three categories of acceleration are arranged in the respective columns. The rows now show the lagged composites of 500 hPa geopotential height (thick contours) and anomalies (color shaded). In each panel, the corresponding composite 1000 hPa geopotential height anomalies are shown by purple contours.

The ensemble average for rapid tangential acceleration cases (left column of Fig. \ref{fig:lagCompZ}) shows a TC moving rapidly towards a extratropical wavepacket. At day-1, the TC circulation is relatively symmetric as depicted by the contours of 1000 hPa geopotential height anomalies. At 500 hPa, the upstream trough is still at a distance while the downstream ridge has linked with the TC outflow. By Day+1, the composite storm has moved further poleward and eastward and is now located between the upper-level upstream trough and downstream ridge in a position that is optimal for surface baroclinic development. The 1000 hPa geopotential field is now asymmetric with a characteristic signal of a cold front.  Over the course of three days, at 500 hPa, the upstream trough as well as the downstream ridge-trough couplet appear to amplify while simultaneously propagating downstream. The 500 hPa geopotential gradient poleward of the storm is also amplified, indicating a strengthening jet streak. These developments are generally consistent with the process of ET. As summarized in \citet{K2019}, several mechanisms are at play: The poleward moving TC perturbs the extratropical flow and excites a developing Rossby wavepacket that disperses energy downstream. Furthermore, the outflow of the TC is a source of low potential vorticity (PV) air that further reinforces the downstream ridge. Overall, it is evident that the ensemble average evolution of rapid acceleration cases resembles the canonical ET process.

On the other hand, for the near zero acceleration cases (middle column Fig. \ref{fig:lagCompZ}), the composite storm remains equatorward of the extratropical wavepacket and maintains a nearly symmetric structure. Furthermore, the extratropical wavepacket shows little change in amplitude and the constituent troughs and ridges remain nearly stationary, with little evidence of downstream development.  Thus, the near zero acceleration cases are not likely to be involved in ET. The rapid deceleration composite (right column Fig. \ref{fig:lagCompZ}) also shows that the extratropical wavepacket is poleward of the TC. However, by Day+1, there is evidence of amplification of the downstream ridge-trough couplet indicating development in the extratropical stormtrack related to the interaction with the TC.  This indicates a likelihood of ET at a later time.

\begin{figure*}[t]
  \centering
    \includegraphics[width=.95\textwidth, trim={0cm 8cm 0cm 4cm},clip]{composite_tang_40_30_NA_500hPa.pdf}
  \caption{}
  \label{fig:lagCompZ}
\end{figure*}
%-------------------------------------------------------

%-------------------------------------------------------
\section{Extratropical Transition}

In this section, we further examine the relationship between TC acceleration and ET.

Fig. \ref{fig:AS} shows the temporal evolution of TC accelerations and speed relative to the time of ET. This was calculated by identifying all TCs that underwent ET and assigning hour 0 to the first instance of ET designation in the track. The figure shows the ensemble average over all

\subsection{Trends}

>>>>

\title{Ensemble Averaged Patterns and Long-term Trends of Acceleration of Tropical Cyclones -- A Proxy For Extratropical Interactions}

@@

\begin{abstract}
  It is well known that rapid changes in tropical cyclone (TC) motion occur during interaction with extratropical wavetrains. While the forward speed of TCs has received much attention in the past, acceleration has not. Using a large data sample, we formally examined the synoptic-scale patterns associated with  tangential and normal components of TC acceleration. During rapid acceleration, a developing wavepacket in the extratropics is present and the TC ?????? with the TC straddled by an upstream trough and downstream ridge. This pattern is consistent with the process of extratropical transition (ET) of TCs. In contrast, during near-zero or rapid deceleration, the extratropical wavepacket remains poleward, with a ridge  located directly north of the storm center.  On average, the tangential acceleration peaks about 18 hours prior to the ET designation. The curvature acceleration increases before ET and stabilizes thereafter.
%
 We also find a robust decrease in the annually averaged values of rapid acceleration over the past 38 years within the latitude band most commonly associated with ET over the Atlantic. Consequently, we hypothesize that trends in rapid acceleration can be used as an independent metric to assess changes in the extratropical stormtrack activity that mediate ET.
\end{abstract}

@@

\introduction
The motion of a tropical cyclone (TC), to a leading order, is known to be  governed by the large-scale environment surrounding it \citep[e.g.][]{Hodanish1993}. It was recognized early on that the forward speed of a TC can be closely approximated by the surrounding wind field \citep{E2018}.  It is also understood that a TC is not an isolated vortex that is passively carried by a current. The background environment also has attendant gradients of potential vorticity, moisture, and wind shear. A TC actively responds to these external stimuli. Consequently, this impacts not only the motion and intensity of the TC, but also the the large scale flow. For instance, the generation of the $\beta-gyres$, which infuences TC motion, is a response of the TC circulation to the background potential vorticity gradient. Similarly, vertical wind shear can modulate the intensity of the TC via ventilation of low entropy air into the storm core. This, in turn,  can also impact subsequent TC motion. On the other hand, a TC can influence the large-scale flow by exciting  waves in the extratropical stormtrack, leading to rapid downstream development and rearrangement of the flow. Thus, it can be contended that a TC is always interacting with its environment and the interaction is, in part, manifested in its motion.

The aspects of TC motion that are particularly important for operational forecasts are track and forward speed. The track garners much attention for obvious reasons -- it  informs  potential locations that will be affected by a storm. TC forward speed is an important parameter that impacts storm surge, intensity changes and precipitation. There is a large body of published literature dealing with various research and operational problems related to TC tracks and speed. They are extensively documented in several recent reviews \citep[e.g.,][]{E2018}. The focus of this paper is the acceleration and deceleration of TCs. While Forecasters have long known that the TCs often speed up when approaching an extratropical wavetrain, relatively less attention appears to have been devoted in the research literature to details of TC acceleration.

When TCs move poleward, they encroach upon the extartropical stormtrack. This leads to an interaction between the TCs and the baroclinic eddies of the stormtrack. The outcome of this interaction is varied. Some TCs weaken and dissipate while others may strengthen and retain their tropical nature for some more time. In the Atlantic, approximately ?\% of storms undergo  extratropical transitoin (ET) - a complex process that

%As noted by \cite{E2018}, steady progress has been made in our understanding of TC motion and this is reflected in our ability to forecast their tracks.

@@

\section{Data}
We used the IBTrACS v4 \citep{KnappIBTRaCS} data for track information. We did not apply a threshold for maximum sustained winds, but discarded all storms that lasted less than three days. Thus our definition of a TC in this paper includes a few tropical depressions that lasted greater than 3 days.

{\bf Cite Kossin and the other paper here.}

We focused on the years 1966--2019, spanning the \emph{satellite era} for the North Atlantic and Pacific basins (e.g.. Beven et al. 2003; Landea 2007; Vecchi and Knutson), for which the TC data is deemed relatively more reliable as compared to previous years. Selected atmospheric fields were taken from the the European Center for Medium Range Weather Forecasting's (ECMWF) ERA-Interim (ERAi) reanalysis \citep{DUS11} for the period 1981--2016.

 As our intent here was to examine the motion of tropical storms, in many of our analysis, we excluded instances when a given storm became extratropical. To ascertain whether a tropical storm underwent ET, we used the \emph{nature} designation in the IBTraCS database that relies on the judgement of the forecasters from one or more agencies responsible for the ocean basin.  Admittedly, there will be some subjectivity in this designation. An alternative would be to employ a metric such as the cyclone phase space \citep{Hart2003}, usually calculated using meteorological fields from global numerical weather prediction models (e.g., reanalysis). As we wish to be independent of modeled products to characterize the storm, we choose to use the forecaster designated storm nature. Furthermore, \cite{BCSEH2019a} found that the phase space calculation was sensitive to the choice of the reanalysis model. Occasionally, the nature of the storm is not recorded (NR) or, if there is an inconsistency among the agencies, it is designated as mixed (MX). In the north Atlantic, only a small fraction of track data ($\approx 0.5\%$) in the IBTrACs is designated as NR or MX. However, the fraction is higher in the North Pacific ($\approx 14\%$). As such, for this basin, we include all storm positions, irrespective of nature designation. We return to this point in a subsequent sections where it could introduce some uncertainty in our conclusion for this basin. For this reason, we focus on Atlantic TCs in this paper, and provide selected results for western North Pacific storms in Appendix B.

\section{Tangential and curvature acceleration}
The acceleration of a hypothetical fluid element moving with the center of a TC can be written as:
%
\begin{equation}
    \mathbf{a} = \frac{dV}{dt} \mathbf{\hat{s}} + \frac{V^2}{R} \mathbf{\hat{n}} 
    \label{eq:acc}
\end{equation}
%
where $V$ is the forward speed and $R$ is the radius of curvature of the track at a given location. Here, $\mathbf{\hat{s}}$ and $\mathbf{\hat{n}}$ are orthogonal unit vectors in the so-called \emph{natural coordinate} system. The former is directed along the TC motion. The latter is directed along the radius of curvature of the track.  The first term in eq. \ref{eq:acc} is the tangential acceleration and the second is the curvature or normal acceleration. The speed $V_j$ at any track location (given by index j)  is calculated as:
%
\begin{equation}
V_j = \frac{(D_{j,j-1} +  D_{j,j+1})}{\delta t}
\end{equation}
%
where $D$ refers to the distance between the two consecutive points, indexed as shown above, along the track. Since we used 3-hourly reports from the IBtracs, $\delta t = 6$ h. The tangential acceleration was calculated from the speed using centered differences.

To calculate the curvature acceleration, it is necessary to first determine the radius of curvature, R. A standard approach to calculating R, given a set of discrete points along a curve (in our case, a TC track), is to fit a circle through three consecutive points. For a curved line on a sphere, it can be shown that:

\begin{subequations}
\renewcommand{\theequation}{\theparentequation.\arabic{equation}}
\begin{equation}
R  = R_e sin^{-1} {\left(\sqrt{\frac{2 d_{12} d_{13} d_{23}} {(d_{12} +d_{13} +d_{23})^2 -2(d_{12}^2 +d_{13}^2 +d_{23}^2)}}\right)}
\end{equation}
%
where $R$ is the radius of curvature, $R_e$ is the radius of the Earth, and the $d$ terms are expressed as follows:
%
\begin{gather}
 d_{12} = 1-(\cos{T_1}\cos{T_2}\cos{(N_2-N_1)} + \sin{T_1}\sin{T_2}) \\
 d_{13} = 1-(\cos{T_1}\cos{T_3}\cos{(N_3-N_1)} + \sin{T_1}\sin{T_3}) \\
 d_{23} = 1-(\cos{T_2}\cos{T_3}\cos{(N_3-N_2)} + \sin{T_2}\sin{T_3})
\end{gather}
\end{subequations}
%
where, $T_1$, $T_2$, and $T_3$ are the latitudes of the 3 points while $N_1$, $N_2$, and $N_3$ are the longitudes. The center of the circle is given by the coordinates:
%
\begin{subequations}
\renewcommand{\theequation}{\theparentequation.\arabic{equation}}
\begin{gather}
\tan{\mathbf{T}} = \pm {\frac{\cos{T_1}\cos{T_2}\sin{(N_2-N_1)} + \cos{T_1}\cos{T_3}\sin{(N_1-N_3)} + \cos{T_2}\cos{T_3}\sin{(N_3-N_2)}} {\sqrt{\alpha^2+\beta^2}}} \\ \tan{\mathbf{N}} = -{\frac{\alpha}{\beta}}
\end{gather}
%
Where $\mathbf{T}$ and $\mathbf{N}$ are the latitude and longitude, respectively of the circle's center, and $\alpha$ and $\beta$ are obtained using:

\begin{gather}
\alpha = \cos{T_1}(\sin{T_2}-\sin{T_3})\cos{N_1} + \cos{T_2}(\sin{T_3} -\sin{T_1})\cos{N_2} +\cos{T_3}(\sin{T_1}-\sin{T_2})\cos{N_3} \\
\beta = \cos{T_1}(\sin{T_2}-\sin{T_3})\sin{N_1} + \cos{T_2}(\sin{T_3}-\sin{T_1}) \sin{N_2} +\cos{T_3}(\sin{T_1}-\sin{T_2})\sin{N_3}
\end{gather}
\end{subequations}

While the expressions above were derived independently for this work, we make no claim of originality since they are based on elementary principles of geometry. Figure \ref{fig:track} shows some examples of the radius of curvature calculation for hurricanes Katrina (2005) and Irma (2017).

@@

During 1966--2019, we identified 689 storms in the North Atlantic that met the 2-day threshold. Figure \ref{fig:distNA} shows some basic statistics for speed and accelerations as a function of latitude for Atlantic storms. TC locations were sorted in $10^\circ$-wide latitude bins for this and all instances classified as NR or ET were excluded. Table \ref{tab:distNA} provides the same data.

The average speed of all North Atlantic TCs (including over-land) is about 21 km/hr. As expected, TC speed clearly varies with latitude. It is lower in the subtropics as compared to other latitudes. It increases sharply in the vicinity of $40^{o}$ N. The tangential acceleration, as defined in eq. \ref{eq:acc}, can be positive or negative. The mean and median tangential acceleration remains near-zero equatorward of $30^\circ$N. This suggests that TCs in this region tend to translate steadily, or are equally likely to accelerate and decelerate. The tangential acceleration is substantially positive poleward of $30^\circ$N. TCs in these latitudes are subject to rapid acceleration. For example, the mean tangential acceleration in the $30^\circ-40^\circ$N latitude band is about 5.0  km/hr day$^{-1}$.  The curvature acceleration, by our definition, takes only positive values and, steadily increases with latitude. The distributions of TC speed and tangential acceleration are relatively symmetric about the median value as compared to the curvature acceleration.

The corresponding information for western North Pacific for 1732 storms that met the 2-day threshold is presented in the appendix (Figure \ref{fig:distWP} and Table \ref{tab:distWP}). In this case, we consider all track points, including NR and ET for reasons discussed in section 2. The basic distribution of speed and accelerations are similar to the North Atlantic except that the storms, on average, were found to have negative tangential acceleration poleward of 40$^\circ$N.

@@

We now examine the flow pattern associated with extremes in tangential and curvature acceleration of tropical storms. For this, storm-relative ensemble average fields are constructed uisng ERA-interim gridded reanalysis over the period 1980--2015. We used the following method.

\begin{itemize}
    \item All TC track locations were binned into 10$^\circ$ wide latitude strips (e.g.,  10-20$^\circ$ N, 20-30$^\circ$ N, 30-40$^\circ$ N). Instances where storms were classified as ET or NR were excluded. For brevity, only results for the 30-40$^\circ$ N bin are discussed here. A total of 
    3480 track points were identified for this latitude bin. Note that a particular TC could appear more than once in a latitude bin at different times.

    \item The tangential and curvature acceleration of all TCs within each latitude bin were sorted in an ascending order.

    \item The accelerations in each bin were separated into two categories based on whether they fell within: (a) The top 90th percentile of all sorted values; or (b) The bottom 10th percentile of all sorted values.

    \item For tangential acceleration, with both positive and negative values, we interpret these two categories as rapid acceleration and rapid deceleration respectively.

    \item For curvature acceleration, with only positive values, we interpret these two categories as rapid acceleration and near-zero acceleration respectively.

    \item For each category, we computed an ensemble average composite field of the geopotential field at selected levels. The composites were calculated after shifting the grids such that the center of all storms were coincident. The centroid position of storms was used for the composite storm center, and the corresponding time was denoted as Day-0. Lag-composites were created by shifting the dates backward and forward relative to Day-0.

    \item For anomaly fields, we subtracted long-term synoptic climatology form the total field. The climatology was calculated for each day of the year by averaging data for that day over the years 1980-2015. This was followed by a 7-day running mean smoother. To account for diurnal fluctuations, the daily climatology calculated for each available synoptic hour in the ERA-interim data (00, 06, 12 and 18 UTC).

    \item Statistical significance of anomaly fields shown in this section were evaluated by comparing them against 1000 composites created by randomly drawing dates from the period July--October, 1980--2015. The sample size for each composite was 350 for the 30-40$^\circ$ N bin. A two-tailed significance was evaluated at the 95\% confidence level with the null hypothesis being that the anomalies could have resulted from a random draw.

@@

\subsubsection{Day 0 (reference day) composite}

Figure \ref{fig:tangComp} shows  storm-centered composite geopotential heights (thick orange lines) and their anomalies (color shaded) for all Atlantic TCs located within 30-40$^\circ$ N. Two categories of tangential acceleration are shown: rapid acceleration (left column) and rapid deceleration (right column).   The  fields are shown at three levels -- 300 hPa, 500 hPa and 850 hPa. In each panel, the anomalous 1000 hPa geopotential is shown using thinner black contours. It highlights the composite TC and the surface development within the extratropical stormtrack.

The ensemble average for rapid tangential acceleration  (left column of Fig. \ref{fig:tangComp}) shows the composite TC interacting with a well defined extratropical wavepacket. The TC is straddled by an upstream trough and a downstream ridge. At 500 hPa, the geopotential anomalies of the TC and the upstream trough are close and, consequently, appear to be connected. This yields a negative tilt in the horizontal and indicates the onset of cyclonic wrap-up of the trough around the TC. The 1000-hPa geopotential anomaly field is dominated by the composite TC. It also shows the relatively weaker near-surface cyclones and anticyclones of the extratropical stormtrack. The entire wavepacket shows upshear tilt of geopotential anomalies with height, indicating baroclinic growth. This arrangement of the TC and the extratropical wavepacket is consistent with the synoptic-scale flow that is typically associated with ET \citep[e.g.,][]{KHE2000,MGY2003,RJD2008,RJ2014,K2019}. At this point, all storms in the ensemble were still classified as \emph{tropical}. Thus, we interpret this composite as pre-ET completion state.

The ensemble average for rapid tangential deceleration cases (right column of Fig. \ref{fig:tangComp}) shows an entirely different synoptic-scale pattern. The extratropical wavepacket is substantially poleward, with a ridge immediately north of the composite TC. The geopotential anomalies of the extratropical wavepacket and the composite TC appear to be distinct at all three levels, with no evidence of merger. The prominent synoptic structure is the cyclone-anticyclone dipole formed by the TC and the extratropical ridge.

%------------------------------------------------------------------------------------------
%

\subsubsection{Lag Composites}
To get a sense of the temporal evolution of the entire system, we show lag composites for Day-2 to Day+2 in Fig. \ref{fig:lagtang}. As in the previous figure, the two categories of acceleration are arranged in the respective columns. The rows now show  500-hPa geopotential height (thick contours) and anomalies (color shaded). In each panel, the corresponding 1000-hPa geopotential height anomalies are shown by thin black contours.

The ensemble average for rapid tangential acceleration (left column of Fig. \ref{fig:lagtang}) shows a TC moving rapidly towards an extratropical wavepacket. At day-2, the TC circulation is relatively symmetric as depicted by the contours of 1000 hPa geopotential anomalies. The downstream extratropical ridge is prominent, but the upstream trough is much weaker at this time. On Day-1, the entire extratropical wavepacket has amplified and the 500 hPa geopotential anomalies of the TC and a developing upstream trough have merged. This process continues through Day 0. By Day+1, the composite storm has moved further poleward and eastward and is now located between the upper-level upstream trough and downstream ridge in a position that is optimal for further baroclinic development. The 1000 hPa geopotential field is now asymmetric with a characteristic signal of a cold front.

The picture that is evident from these 500-hPa composite fields is that, over the course of the 4 days, downstream ridge-trough couplet amplifies while simultaneously propagating eastward. The upstream trough cyclonically wraps around the TC and the two have merged by Day 1. The geopotential gradient poleward of the storm is also enhanced, indicating a strengthening jet streak. These features are consistent the process of ET \citep[e.g.,][]{K2019}. The poleward moving TC may either interact with an existing wavepacket or perturb the extratropical flow and excite a Rossby wavepacket that disperses energy downstream \citep[e.g.,][]{RJ2014}. The outflow of the TC is a source of low potential vorticity (PV) air that further reinforces the downstream ridge \citep[e.g.,][]{RJD2008}. To further illustrate the interaction, the tracks of the TC and the 500-hPa ridge of the extratropical wavepacket  are presented in Fig. \ref{fig:tang_track}a. It can be clearly seen that the TC merges with the  extratropical stormtrack. Furthermore, the 500 hPa ridge has rapidly moved downstream during the 4 day period, indicating a very progressive pattern. The eastward phase speed of the extratropical wavepacket,  as inferred from the track of the ridge, is $\approx 7$ ms$^{-1}$. The TC speed, averaged over the same 4-day period,  is $\approx 6$ ms$^{-1}$. The close correspondence between the two and the  merger of the tracks further supports the notion that the synoptic-scale evolution during rapid acceleration cases is consistent with the canonical pattern associated with ET.

On the other hand, during rapid deceleration  (right column Fig. \ref{fig:lagtang}), the composite TC remains equatorward of the extratropical wavepacket and maintains a nearly symmetric structure throughout the period.  At all times, the arrangement of the TC and the extratropical ridge directly poleward is akin to a vortex dipole. The extratropical wavepacket is not as progressive as in the rapid acceleration case. This is seen clearly from the tracks of the TC and the ridge (Fig. \ref{fig:tang_track}b). The phase speed of the extratropical wavepacket is $\approx 3$ ms$^{-1}$ while the TC speed is $\approx 1.5$ ms$^{-1}$.  The phasing of the TC and the extratropical wavepacket has led to the formation of a cyclone-anticyclone vortex dipole. While this structure is not consistent with a typical ET process, the TC and the extratropical wavepacket are certainly interacting. If viewed from a potential vorticity (PV) perspective, the induced flow from the cyclonic vortex (TC) will be westward within the poleward anticyclonic vortex (ridge). Similarly, the induced flow from the ridge will also be westward at the location of the TC. The combined effect will be a mutual slowdown as reflected in the speeds estimated from Fig. \ref{fig:tang_track}b.

%wavepacket remain nearly stationary, with little evidence of downstream development.  Thus, the near zero acceleration cases are not likely to be involved in ET. The rapid deceleration composite (right column Fig. \ref{fig:lagCompZ}) also shows that the extratropical wavepacket is poleward of the TC. However, by Day+1, there is evidence of amplification of the downstream ridge-trough couplet indicating development in the extratropical stormtrack related to the interaction with the TC.  This indicates a likelihood of ET at a later time.

\subsection{Curvature Acceleration}

\subsubsection{Day 0 (reference day) composite}

As in the  previous section, Figure \ref{fig:curvComp} shows storm-centered ensemble averages for two categories of curvature acceleration. The composite for rapid acceleration (left column) shows a TC that is primarily interacting with an extratropical ridge that is poleward and downstream of it. The upstream trough in the extratropics is weaker and farther westward as compared with the rapid tangential acceleration (Fig. \ref{fig:curvComp}a). Furthermore, instead of the upstream trough wrapping cyclonically, in this case,  we see the downstream ridge wrapping anticyclonically around the TC. The composite for near-zero curvature acceleration (right column) is quite similar to the composite for rapid tangential deceleration (Fig. \ref{fig:curvComp}d--f). The extratropical wavepacket is poleward and the TC-ridge system appears as a vortex dipole.

\subsubsection{Lag Composites}
The temporal evolution of the entire system for the two categories of curvature acceleration is shown in Fig. \ref{fig:lagcurv}. For rapid curvature acceleration, we see a TC that is moving poleward towards an extratropical ridge. During the subsequent days, the ridge moves eastward initially and begins to wrap around the TC. This arrangement promotes the curving of the TC. By Day+2, the anticyclonic wrapping and thinning has resulted in a significantly weaker ridge as compared to a few days prior. For the near-zero curvature cases ( Fig. \ref{fig:lagcurv}f--g), the initial movement of the TC is also directly poleward towards the extratropical ridge. However, in this case, the ridge remains poleward of the TC. There is also significantly less anticyclonic wrapping of ridge. The TC-ridge system takes the form of a cyclonic-anticyclonic vortex pair similar to the rapid tangential deceleration composite.
